Reproduction frequency and offspring survival decline in elephant seals past prime age

### 2024-07-16 paper/paperv3.qmd
model clarity and TODOs

resight <- read.csv(here::here("data/raw/fullresights.csv")) %>% 
  select(animalID, resightID, season, date, timeofyear, calyear, yday) %>% 
  mutate(animalID = as.factor(animalID))

@@

ggplot(breed_pred_wgt, aes(age, predicted)) +
  # Predictions for individual years
  geom_line(aes(group = interaction(age_cat, year_fct)),
            breed_pred_years,
            alpha = 0.1) +
  # Weighted model (CI and mean response)
  geom_ribbon(aes(fill = age_cat, ymin = conf_lo, ymax = conf_hi), 
              alpha = 0.5) +
  geom_line(aes(color = age_cat),
            linewidth = 1.2) +
  # Population-level mean response
  geom_line(aes(y = predicted_pop, group = age_cat),
            linetype = "dotted",
            linewidth = 1.2,
            alpha = 0.5) +
  # Summarized raw data
  geom_pointrange(aes(y = perc_breed,
                      ymin = breed_lwr,
                      ymax = breed_upr),
                  breed_nonbreed) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_y_continuous("Breeding probability", 
                     labels = scales::percent,
                     limits = c(NA, 1.02),
                     breaks = seq(0.4, 1.0, by = 0.2)) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  labs(x = "Age (years)") +
  theme(legend.position = "none") + 
  geom_text(aes(label = n), y = 1, vjust = -0.5)

@@

fig_surv <- ggplot(surv_pred_wgt, aes(age, predicted)) +
  # Individual years
  geom_line(aes(group = interaction(age_cat, year_fct)),
            surv_pred_years,
            alpha = 0.1) +
  # Weighted model (confidence interval and mean response)
  geom_ribbon(aes(fill = age_cat, ymin = conf_lo, ymax = conf_hi), 
              alpha = 0.2) +
  geom_line(aes(color = age_cat, linetype = age_cat), linewidth = 1.2) +
  # Population-level mean response
  geom_line(aes(y = predicted_pop, group = age_cat),
            linetype = "dotted", linewidth = 1.2, alpha = 0.5) +
  # Raw data
  geom_pointrange(aes(y = perc_surv, 
                      ymin = surv_lwr, 
                      ymax = surv_upr),
                  surv_summ) +
  # Senescence threshold
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_y_continuous("Offspring survival", 
                     labels = scales::percent,
                     limits = c(0, 1)) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  scale_linetype_manual(values = c("dashed", "solid")) +
  labs(x = "Age (years)") +
  theme(legend.position = "none") +
  geom_text(aes(label = n), y = 1, vjust = 1)

@@

fig_recr <- ggplot(recr_pred, aes(age, predicted)) +
  # Individual years
  geom_line(aes(group = interaction(age_cat, year_fct)),
            recr_pred_years,
            alpha = 0.1) +
  # Weighted model (confidence interval and mean response)
  geom_ribbon(aes(fill = age_cat, ymin = conf_lo, ymax = conf_hi), 
              alpha = 0.2) +
  geom_line(aes(color = age_cat), linetype = "dashed", linewidth = 1.2) +
  # Population-level mean response
  geom_line(aes(y = predicted_pop, group = age_cat),
            linetype = "dotted", linewidth = 1.2, alpha = 0.5) +
  # Raw data
  geom_pointrange(aes(y = perc_recr, 
                      ymin = recr_lwr, 
                      ymax = recr_upr),
                  recr_summ) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_y_continuous("Offspring recruitment", 
                     labels = scales::percent,
                     limits = c(0, 1)) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  labs(x = "Age (years)") +
  theme(legend.position = "none") + 
  geom_text(aes(label = n), y = 1, vjust = 1)

@@

$$
\begin{align}
d &\sim NB(\mu, k) \\
\mu &= exp(\eta) \\
\eta &= \beta_0 + \beta_p * a + Zy
\end{align}
$$

Where: $d$ is the duration of the phase in days, $NB$ is the negative binomial distribution with location $\mu$ and dispersion $k$, $\eta$ is the linear predictor, $\beta_0$ is the intercept, $\beta_p$ is the coefficient for age ($a$) by phase (i.e., for $p$ in "molting" or "breeding"), $y$ is the year, and $Z$ is the design matrix relating $y$ to $\eta$.

@@

breed_duration_plot <- ggpredict(
  pheno_model, 
  terms = c("age10 [all]", "phase", "year [all]")
) %>% 
  as_tibble() %>% 
  transmute(age = x * 10,
            dur_days = predicted,
            dur_days_lwr = conf.low,
            dur_days_upr = conf.high,
            phase = group,
            year = as.numeric(as.character(facet))) %>% 
  left_join(n_phenology, by = "year") %>% 
  group_by(age) %>% 
  filter(phase == "breeddur") %>% 
  summarize(across(starts_with("dur_days"), \(x) weighted.mean(x, n)),
            .groups = "drop") %>% 
  ggplot(aes(age)) +
  geom_line(aes(age, predicted, group = interaction(phase, facet)), 
            phen_pred_years, 
            alpha = 0.1) + 
  geom_ribbon(aes(ymin = dur_days_lwr, ymax = dur_days_upr),
              alpha = 0.25, 
              fill = "#D05794") +
  geom_pointrange(aes(y = dur_days, 
                      ymin = dur_days_lwr, 
                      ymax = dur_days_upr),
                  breed_pheno_raw,
                  position = position_dodge(width = 0.5), 
                  color = "#D05794") +
  geom_line(aes(y = dur_days), 
            linetype = "dashed", 
            color = "#D05794") +
  labs(x = "Maternal age (years)",
       y = "Breeding haul out (days)") +
  theme(legend.position = "none",
        axis.text.y = element_text(size = 8))

molt_duration_plot <- ggpredict(
  pheno_model,
  terms = c("age10 [all]", "phase", "year [all]")
) %>%
  as_tibble() %>%
  transmute(age = x * 10,
            dur_days = predicted,
            dur_days_lwr = conf.low,
            dur_days_upr = conf.high,
            phase = group,
            year = as.numeric(as.character(facet))) %>%
  left_join(n_phenology, by = "year") %>%
  group_by(age, phase) %>%
  filter(phase == "moltdur") %>%
  summarize(across(starts_with("dur_days"), \(x) weighted.mean(x, n)),
            .groups = "drop") %>%
  ggplot(aes(age)) +
  geom_line(aes(age, predicted, group = interaction(phase, facet)), 
            phen_pred_years, 
            alpha = 0.1) + 
  geom_ribbon(aes(ymin = dur_days_lwr, ymax = dur_days_upr),
              alpha = 0.25, 
              fill = "#E09A2D") +
  geom_pointrange(aes(y = dur_days, 
                      ymin = dur_days_lwr, 
                      ymax = dur_days_upr),
                  molt_pheno_raw,
                  position = position_dodge(width = 0.5), 
                  color = "#E09A2D") +
  geom_line(aes(y = dur_days), 
            color = "#E09A2D") +
  labs(x = "Maternal age (years)",
       y = "Molting haul out (days)") +
  theme(legend.position = "none", 
        axis.text.y = element_text(size = 8))

@@

# Are hypotheses supported by the model?
extract_p <- \(h) map_dbl(mod_sens, \(m) pnorm(m[[h]]["age10:age_catOld", "z value"]))
p_tbl <- tibble(age_cutoff = cutoff, 
                "H1 (breeding)" = extract_p("h1"),
                "H2a (survival)" = extract_p("h2_surv"),
                "H2b (recruitment)" = extract_p("h2_recr"),
                "H3 (sex ratio)" = extract_p("h3"))

@@

# Biological age model
breed_bioage_mod <- glmer(
  observed_int ~ bioage10 : bioage_cat + (1 | animalID) + (1 | year_fct),
  sealdat2,
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

@@

# Breakpoint comparison
p_by_tau <- map_dbl(-12:-3, \(tau) {
  mod <- sealdat %>%
    drop_na(lastobs) %>%
    mutate(bioage = age - lastobs,
           bioage10 = (bioage - tau) / 10,
           bioage_cat = factor(bioage < tau,
                               labels = c("Old", "Young"))) %>% 
    glmer(
      observed_int ~ bioage10 : bioage_cat + (1 | animalID) + (1 | year_fct),
      .,
      family = "binomial",
      control = glmerControl(optimizer = "bobyqa")
    )
  old_z <- summary(mod)$coef["bioage10:bioage_catOld", "z value"] 
  pnorm(old_z)
})

@@

# Model comparison
breed_age_mod2 <- glmer(
  observed_int ~ age10 : age_cat + (1 | animalID) + (1 | year_fct),
  drop_na(sealdat2, lastobs),
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

@@

# Summarize the raw data
bioage_summ <- sealdat2 %>% 
  group_by(bioage) %>% 
  summarize(n_obs = n(), 
            n_breed = sum(observed_int),
            perc_breed = n_breed / n_obs, 
            breed_lwr = binom.test(n_breed, n_obs)$conf.int[1], 
            breed_upr = binom.test(n_breed, n_obs)$conf.int[2])

# Create ANNUAL (i.e. rand effect of year only) predictions
bioage_pred_years <- ggpredict(
  breed_bioage_mod, 
  terms = c("bioage10 [all]", "bioage_cat", "year_fct"),
  type = "random"
) %>% 
  as_tibble() %>% 
  mutate(bioage = x * 10 + bioage_senesce,
         bioage_cat = factor(group, levels = c("Young", "Old")),
         year_fct = factor(facet)) %>% 
  filter((bioage_cat == "Young" & bioage < bioage_senesce) |
           (bioage_cat == "Old" & bioage >= bioage_senesce),
         bioage <= 20)

# Create WEIGHTED-AVERAGE predictions using years' random effects
# ggpredict can't do that for confidence intervals
# create prediction grid
bioage_grid <- expand_grid(
  bioage = -17:0,
  year_fct = factor(1988:2019)
) %>% 
  mutate(bioage10 = (bioage - bioage_senesce) / 10,
         bioage_cat = factor(ifelse(bioage < bioage_senesce, "Young", "Old"),
                          levels = c("Young", "Old")))

@@

@@

@@

>>>>

resight <- read_csv(here::here("data/raw/fullresights.csv"), 
                    show_col_types = FALSE) %>% 
  select(animalID, resightID, season, date, timeofyear, calyear, yday) %>% 
  mutate(animalID = as.factor(animalID))

@@

ggplot(breed_pred_wgt, aes(age, predicted)) +
  # Predictions for individual years
  geom_line(aes(group = interaction(age_cat, year_fct)),
            breed_pred_years,
            alpha = 0.1) +
  # Weighted model (CI and mean response)
  geom_ribbon(aes(fill = age_cat, ymin = conf_lo, ymax = conf_hi), 
              alpha = 0.5) +
  geom_line(aes(color = age_cat),
            linewidth = 1.2) +
  # Population-level mean response
  geom_line(aes(y = predicted_pop, group = age_cat),
            linetype = "dotted",
            linewidth = 1.2,
            alpha = 0.5) +
  # Summarized raw data
  geom_pointrange(aes(y = perc_breed,
                      ymin = breed_lwr,
                      ymax = breed_upr),
                  breed_nonbreed) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_y_continuous("Breeding probability", 
                     labels = scales::percent,
                     breaks = seq(0.4, 1.0, by = 0.1)) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  coord_cartesian(ylim = c(0.68, 1.02)) +
  labs(x = "Age (years)") +
  theme(legend.position = "none") + 
  geom_text(aes(label = n), y = 1, vjust = -0.5)

@@

fig_surv <- ggplot(surv_pred_wgt, aes(age, predicted)) +
  # Individual years
  geom_line(aes(group = interaction(age_cat, year_fct)),
            surv_pred_years,
            alpha = 0.1) +
  # Weighted model (confidence interval and mean response)
  geom_ribbon(aes(fill = age_cat, ymin = conf_lo, ymax = conf_hi), 
              alpha = 0.2) +
  geom_line(aes(color = age_cat, linetype = age_cat), linewidth = 1.2) +
  # Population-level mean response
  geom_line(aes(y = predicted_pop, group = age_cat),
            linetype = "dotted", linewidth = 1.2, alpha = 0.5) +
  # Raw data
  geom_pointrange(aes(y = perc_surv, 
                      ymin = surv_lwr, 
                      ymax = surv_upr),
                  surv_summ) +
  # Senescence threshold
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_y_continuous("Offspring survival", 
                     labels = scales::percent) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  scale_linetype_manual(values = c("dashed", "solid")) +
  coord_cartesian(ylim = c(0, 0.6)) +
  labs(x = "Age (years)") +
  theme(legend.position = "none") + 
  geom_label(aes(label = n), y = 0.6, vjust = 1)

@@

fig_recr <- ggplot(recr_pred, aes(age, predicted)) +
  # Individual years
  geom_line(aes(group = interaction(age_cat, year_fct)),
            recr_pred_years,
            alpha = 0.1) +
  # Weighted model (confidence interval and mean response)
  geom_ribbon(aes(fill = age_cat, ymin = conf_lo, ymax = conf_hi), 
              alpha = 0.2) +
  geom_line(aes(color = age_cat), linetype = "dashed", linewidth = 1.2) +
  # Population-level mean response
  geom_line(aes(y = predicted_pop, group = age_cat),
            linetype = "dotted", linewidth = 1.2, alpha = 0.5) +
  # Raw data
  geom_pointrange(aes(y = perc_recr, 
                      ymin = recr_lwr, 
                      ymax = recr_upr),
                  recr_summ) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_y_continuous("Offspring recruitment", 
                     labels = scales::percent) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  coord_cartesian(ylim = c(0, 0.6)) +
  labs(x = "Age (years)") +
  theme(legend.position = "none") + 
  geom_label(aes(label = n), y = 0.6, vjust = 1)

@@

$$
\begin{align}
d &\sim NB(\mu, k) \\
e^\mu &= \beta_0 + \beta_p a + Zy
\end{align}
$$

Where: $d$ is the duration of the phase in days, $NB$ is the negative binomial distribution with location $\mu$ and dispersion $k$, $\beta_0$ is the intercept, $\beta_p$ is the coefficient for age ($a$) by phase (i.e., for $p$ in "molting" or "breeding"), $y$ is the year, and $Z$ is the design matrix relating $y$ to $e^\mu$.

@@

breed_duration_plot <- ggpredict(
  pheno_model, 
  terms = c("age10 [all]", "phase", "year [all]")
) %>% 
  as_tibble() %>% 
  transmute(age = x * 10,
            dur_days = predicted,
            dur_days_lwr = conf.low,
            dur_days_upr = conf.high,
            phase = group,
            year = as.numeric(as.character(facet))) %>% 
  left_join(n_phenology, by = "year") %>% 
  group_by(age) %>% 
  filter(phase == "breeddur") %>% 
  summarize(across(starts_with("dur_days"), \(x) weighted.mean(x, n)),
            .groups = "drop") %>% 
  ggplot(aes(age)) +
  geom_line(aes(age, predicted, group = interaction(phase, facet)), 
            filter(phen_pred_years, phase == "breeddur"), 
            alpha = 0.1) + 
  geom_ribbon(aes(ymin = dur_days_lwr, ymax = dur_days_upr),
              alpha = 0.25, 
              fill = "#D05794") +
  geom_pointrange(aes(y = dur_days, 
                      ymin = dur_days_lwr, 
                      ymax = dur_days_upr),
                  breed_pheno_raw,
                  position = position_dodge(width = 0.5), 
                  color = "#D05794") +
  geom_line(aes(y = dur_days), 
            linetype = "dashed", 
            color = "#D05794") +
  labs(x = "Maternal age (years)",
       y = "Breeding haul out (days)") +
  theme(legend.position = "none",
        axis.text.y = element_text(size = 8))

molt_duration_plot <- ggpredict(
  pheno_model,
  terms = c("age10 [all]", "phase", "year [all]")
) %>%
  as_tibble() %>%
  transmute(age = x * 10,
            dur_days = predicted,
            dur_days_lwr = conf.low,
            dur_days_upr = conf.high,
            phase = group,
            year = as.numeric(as.character(facet))) %>%
  left_join(n_phenology, by = "year") %>%
  group_by(age, phase) %>%
  filter(phase == "moltdur") %>%
  summarize(across(starts_with("dur_days"), \(x) weighted.mean(x, n)),
            .groups = "drop") %>%
  ggplot(aes(age)) +
  geom_line(aes(age, predicted, group = interaction(phase, facet)), 
            filter(phen_pred_years, phase == "moltdur"), 
            alpha = 0.1) + 
  geom_ribbon(aes(ymin = dur_days_lwr, ymax = dur_days_upr),
              alpha = 0.25, 
              fill = "#E09A2D") +
  geom_pointrange(aes(y = dur_days, 
                      ymin = dur_days_lwr, 
                      ymax = dur_days_upr),
                  molt_pheno_raw,
                  position = position_dodge(width = 0.5), 
                  color = "#E09A2D") +
  geom_line(aes(y = dur_days), 
            color = "#E09A2D") +
  labs(x = "Maternal age (years)",
       y = "Molting haul out (days)") +
  theme(legend.position = "none", 
        axis.text.y = element_text(size = 8))

@@

# Are hypotheses supported by the model?
extract_p <- \(h) map_dbl(mod_sens, \(m) pnorm(m[[h]]["age10:age_catOld", "z value"]))
p_tbl <- tibble(age_cutoff = cutoff, 
                "H1 (breeding)" = extract_p("h1"),
                "H2a (survival)" = extract_p("h2_surv"),
                "H2b (recruitment)" = extract_p("h2_recr"),
                "H3 (sex ratio)" = extract_p("h3") * 2)

@@

# Biological age model
breed_bioage_mod <- glmer(
  observed_int ~ bioage10 : bioage_cat + (1 | animalID) + (1 | year_fct),
  filter(sealdat2, bioage < 0),
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

@@

# Breakpoint comparison
p_by_tau <- map_dbl(-12:-3, \(tau) {
  mod <- sealdat %>%
    drop_na(lastobs) %>%
    mutate(bioage = age - lastobs,
           bioage10 = (bioage - tau) / 10,
           bioage_cat = factor(bioage < tau,
                               labels = c("Old", "Young"))) %>% 
    filter(bioage < 0) %>% 
    glmer(
      observed_int ~ bioage10 : bioage_cat + (1 | animalID) + (1 | year_fct),
      .,
      family = "binomial",
      control = glmerControl(optimizer = "bobyqa")
    )
  old_z <- summary(mod)$coef["bioage10:bioage_catOld", "z value"] 
  pnorm(old_z)
})

@@

# Model comparison
breed_age_mod2 <- glmer(
  observed_int ~ age10 : age_cat + (1 | animalID) + (1 | year_fct),
  filter(sealdat2, bioage < 0),
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

@@

# Summarize the raw data
bioage_summ <- sealdat2 %>%
  # exclude n=1 age class
  filter(bioage >= -15) %>% 
  group_by(bioage) %>% 
  summarize(n_obs = n(), 
            n_breed = sum(observed_int),
            perc_breed = n_breed / n_obs, 
            breed_lwr = binom.test(n_breed, n_obs)$conf.int[1], 
            breed_upr = binom.test(n_breed, n_obs)$conf.int[2])

# Create ANNUAL (i.e. rand effect of year only) predictions
bioage_pred_years <- ggpredict(
  breed_bioage_mod, 
  terms = c("bioage10 [all]", "bioage_cat", "year_fct"),
  type = "random"
) %>% 
  as_tibble() %>% 
  mutate(bioage = x * 10 + bioage_senesce,
         bioage_cat = factor(group, levels = c("Young", "Old")),
         year_fct = factor(facet)) %>% 
  filter((bioage_cat == "Young" & bioage < bioage_senesce) |
           (bioage_cat == "Old" & bioage >= bioage_senesce),
         bioage >= -15)

# Create WEIGHTED-AVERAGE predictions using years' random effects
# ggpredict can't do that for confidence intervals
# create prediction grid
bioage_grid <- expand_grid(
  bioage = -15:-1,
  year_fct = factor(1988:2018)
) %>% 
  mutate(bioage10 = (bioage - bioage_senesce) / 10,
         bioage_cat = factor(ifelse(bioage < bioage_senesce, "Young", "Old"),
                          levels = c("Young", "Old")))

@@

TODO: fix sample size for bioage=0

@@

TODO: include from age 4 on

@@

TODO: AIC comparison
